### 2018-03-16 paper/4-build.tex
Work on the implementation

\begin{minted}{haskell}
data Traces k v = Traces
    { traces :: [Trace k v]
    , cache  :: Map (Hash v) v }
\end{minted}

We have a list of traces, plus a \hs{Map} from the hash to the actual contents. Checking a trace is the same as before, but if the result is the only thing that is different we can simply retrieve a fresh result from \hs{cache} \emph{without} recomputing it. We split the traces and caches as in cloud interactions to a remote server the checking system may have to examine many traces/hashes, but only at most one complete file.

>>>>

\begin{minted}{haskell}
data Traces k v = Traces
    { traces :: [Trace k v]
    , contents  :: Map (Hash v) v }
\end{minted}

We have a list of traces, plus a \hs{Map} from the hash to the actual contents. Checking a trace is the same as before, but if the result is the only thing that is different we can simply retrieve a fresh result from \hs{contents} \emph{without} recomputing it. We split the traces and contents as in cloud interactions to a remote server the checking system may have to examine many traces/hashes, but only at most one complete file.

### 2018-03-16 paper/5-implementations.tex
Work on the implementation

The helper function \hs{reordering}, whose implementation we omit since it is
technical and not particularly enlightening, calls the function \hs{process} to
\emph{try to build} a \hs{key} by executing the action \hs{act}, in the order
determined by the calculation chain. To decide whether to \hs{rebuild} the
\hs{key}, \hs{process} checks if the \hs{key} itself is marked dirty or the
approximation of its dependencies contains a dirty key. Notice that if the
dependencies of the \hs{key} are \hs{Unknown}, e.g. when it uses the
\cmd{INDIRECT} function, the \hs{key} is always rebuilt. If a \hs{rebuild} is
not needed, \hs{process} returns \hs{Nothing} to indicate this. Otherwise, it
executes \hs{act} leading to two possible \hs{result}s:

@@

The \Shake approach for dependency tracking involves recording traces and verifying them, for which we use the \hs{Trace} type defined in \ref{sec-verifying-traces}. Complete code is given in \S\ref{fig-shake-implementation}.

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Determine whether a trace is relevant to the current state
traceMatch :: (Monad m, Eq k) => (k -> Hash v -> m Bool) -> k -> [Trace k v] -> m [Hash v]
traceMatch check key ts = mapMaybeM f ts
    where f (Trace k dkv v) = do
                b <- return (key == k) &&^ allM (uncurry check) dkv
                return $ if b then Just v else Nothing
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Recursive dependency strategy
recursive
    :: Eq k => (k -> (k -> State (Store i k v, [k]) v) -> State (Store i k v, [k]) (v, [k]) -> State (Store i k v, [k]) ())
    -> Build Monad i k v
recursive step task key store = fst $ execState (ensure key) (store, [])
    where
        ensure key = do
            let fetch k = do ensure k; gets (getValue k . fst)
            s <- get
            when (key `notElem` snd s) $ do
                modify $ \(s, done) -> (s, key:done)
                case trackM task fetch key of
                    Nothing -> return ()
                    Just act -> step key fetch act
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Shake build system
shake :: (Eq k, Hashable v) => Build Monad [Trace k v] k v
shake = recursive $ \key fetch act -> do
    traces <- gets (getInfo . fst)
    poss <- traceMatch (\k v -> (==) v . hash <$> fetch k) key traces
    current <- gets (getHash key . fst)
    when (current `notElem` poss) $ do
        (v, ds) <- act
        modify $ \(s, done) ->
            let t = Trace key [(d, getHash d s) | d <- ds] (getHash key s)
            in (putInfo (t : getInfo s) (putValue key v s), done)

@@

This implementation uses the \hs{recursive} function which requires a function which gets the key to build (\hs{k}), a way to fetch the result of a dependency (\hs{fetch}), and a way to build \hs{k} (\hs{act}). The code first looks up the traces that match the current situation, creating a list of possible results as \hs{poss}, using the helper \hs{traceMatch}. We check if the current value for the key is one of the possible results, and it not, build the key and store it in the trace. We can implement \hs{traceMatch}, which is used in all build systems working with \hs{Trace}, as:

The function \hs{traceMatch} requires a way to check a dependency, and the key, and the traces. We return the \hs{result} field from all traces where the key and dependencies match. We define the recursive combinator as:

The required \hs{step} function gets given the \hs{k} to build, the dependencies as \hs{Maybe [k]} (where they are \hs{Nothing} if the rule is truly monadic), a function to demand the result of a dependent key, and a function to compute the current value and return the actual dependencies. To ensure each file is only visited once in a single execution a piece of temporary state is used with \hs{getTemp} and \hs{modifyTemp} to record which files were visited - this state is not persisted to the \hs{i} info parameter and is reset on each fresh execution of \hs{M}.

@@

Now we have seen the three dependency schemes, we can directly reuse \hs{topological} to define \Bazel. Furthermore, as \Bazel is a tracing build system, we can reuse \hs{Trace} and \hs{traceMatch}, as part of the \hs{Traces} type. We define \Bazel as:

\begin{minted}{haskell}
bazel :: Hashable v => Build Applicative (Traces k v) k v
bazel = topological $ \k dk act -> do
    poss <- traceMatch (\k v -> (== v) <$> getStoreHash k) k . traces =<< getInfo
    if null poss then do
        v <- act
        dh <- mapM getStoreHash dk
        modifyInfo $ \i -> i
            {traces = Trace k (zip dk dh) (getHash v) : traces i
            ,contents = Map.insert (getHash v) v $ contents i}
        putStore k v
    else do
        h <- getStoreHash k
        when (h `notElem` poss) $
            putStore k . (Map.! head poss) . contents =<< getInfo
\end{minted}

>>>>

The helper function \hs{reordering}, whose 21-line implementation we omit since it is
technical and not particularly enlightening, calls the function \hs{process} to
\emph{try to build} a \hs{key} by executing the action \hs{act}, in the order
determined by the calculation chain. To decide whether to \hs{rebuild} the
\hs{key}, \hs{process} checks if the \hs{key} itself is marked dirty or the
approximation of its dependencies contains a dirty key. Notice that if the
dependencies of the \hs{key} are \hs{Unknown}, e.g. when it uses the
\cmd{INDIRECT} function, the \hs{key} is always rebuilt. If a \hs{rebuild} is
not needed, \hs{process} returns \hs{Nothing} to indicate this. Otherwise, it
executes \hs{act} leading to two possible \hs{result}s:

@@

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Determine whether a trace is relevant to the current state
traceMatch :: (Monad m, Eq k) => (k -> Hash v -> m Bool) -> k -> [Trace k v] -> m [Hash v]
traceMatch check key ts = mapMaybeM f ts
    where f (Trace k dkv v) = do
                b <- return (key == k) &&^ allM (uncurry check) dkv
                return $ if b then Just v else Nothing
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Recursive dependency strategy
recursive
    :: Eq k => (k -> (k -> State (Store i k v, [k]) v) -> State (Store i k v, [k]) (v, [k]) -> State (Store i k v, [k]) ())
    -> Build Monad i k v
recursive process task key store = fst $ execState (ensure key) (store, [])
    where
        ensure key = do
            let fetch k = do ensure k; gets (getValue k . fst)
            done <- gets snd
            when (key `notElem` done) $ do
                modify $ \(s, done) -> (s, key:done)
                case trackM task fetch key of
                    Nothing -> return ()
                    Just act -> process key fetch act
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Shake build system
shake :: (Eq k, Hashable v) => Build Monad [Trace k v] k v
shake = recursive $ \key fetch act -> do
    traces <- gets (getInfo . fst)
    poss <- traceMatch (\k v -> (==) v . hash <$> fetch k) key traces
    current <- gets (getHash key . fst)
    when (current `notElem` poss) $ do
        (v, ds) <- act
        modify $ \(s, done) ->
            let t = Trace key [(d, getHash d s) | d <- ds] (getHash key s)
            in (putInfo (t : getInfo s) (putValue key v s), done)

@@

The \Shake approach for dependency tracking involves recording traces and verifying them, for which we use the \hs{Trace} type defined in \ref{sec-verifying-traces}. Complete code is given in \S\ref{fig-shake-implementation}, split into three functions:

\begin{description}
\item[\hs{traceMatch}] takes a list of all recorded \hs{Trace} values, the \hs{key} you are currently building, and a function \hs{check} which checks a specified dependency. It returns the \hs{result} field of all traces that match. Since \hs{check} is in an arbitrary monad, the function has to use \hs{allM}/\hs{&&^} instead of the more usual \hs{all}/\hs{&&} functions\footnote{These functions are all available in the \cmd{extra} library on Hackage.}.
\item[\hs{recursive}] defines the dependency ordering pattern. It requires a \hs{process} function that builds a key given a way to build a dependency, and a way to build the result. The main purpose of \hs{recursive} is to ensure that in a single run any key is not built twice -- for which it extends the \hs{State} monad with a list of \hs{done} keys.
\item[\hs{shake}] ties everything together. First it tests if the traces allow the current state of the target key. If not, it builds the key and updates the store. The only subtlty is that Shake calls \hs{fetch} on the keys while checking them -- building the last-recorded dependencies before checking them.
\end{description}

@@

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
bazel :: (Eq k, Hashable v) => Build Applicative (Traces k v) k v
bazel = topological $ \key ds act -> do
    s <- get
    let Traces traces contents = getInfo s
    poss <- traceMatch (\k v -> return $ getHash k s == v) key traces
    if null poss then do
        v <- act
        modify $ \s ->
            let t = Trace key [(d, getHash d s) | d <- ds] (getHash key s)
                ts = Traces (t : traces) (Map.insert (hash v) v contents)
            in putInfo ts (putValue key v s)
    else do
        when (getHash key s `notElem` poss) $
            modify $ putValue key (contents Map.! head poss)
\end{minted}
\vspace{-2mm}
\caption{An implementation of \Bazel using our framework.}\label{fig-bazel-implementation}
\vspace{-2mm}
\end{figure}

Now we have seen all three dependency schemes, we can reuse , we can directly reuse \hs{topological} to define \Bazel. Furthermore, as \Bazel is a tracing build system, we can reuse \hs{Trace} and \hs{traceMatch}, as part of the \hs{Traces} type. We define \Bazel as:
